The error when inferring phylogenies with incipient species by a Birth-Death model

### 2018-04-09 article.tex
EOD

@@

% Use TODO notes
\usepackage{todonotes}

% Rename 'Abstract' to 'Summary 
\usepackage[english]{babel}
\addto{\captionsenglish}{\renewcommand{\abstractname}{Summary}}

@@

The model prior most relevant here, is the tree prior. The tree prior
embodies the speciation model assumed. Each speciation model has a
different expected branch length distribution. It has been shown 
[NOTE: where?] that the effect of the prior on the posterior distribution 
of trees can be substantial.

@@

There are problems with ignoring protracted speciation. A very pragmatic
example is the conservation of an underestimated amount of species. 
But from a more theoretical perspective, the
observed decline of speciation rate in time (excluding the contemporary human-driven
mass extinction event) may be explained by these new species not being 
recognized yet. Another problem is that the MRCA of young species is estimated 
incorrectly, as in the BD model a species can be ultra-young (as speciation
is assumed to be instantaneous), where the PBD model will place these MRCAs
further back in time (as two events need to have finished completion).

From a theoretical point of view, there is another reason why protracted
speciation has yet become a member of our toolbox: there is no framework
where it fits in: a framework assumes either an analysis at the species or
subspecies level. When picking a tool that does its analysis at the
species level, there is no protocol how to select which incipient species
gets to represent the full species it is still a member of. When picking
a tool at the sub-species level, the incipient state disappears as a
sub-species lineage, and the process degrades to an ordinary birth-death 
process.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l}
    \hline
    $n$ & Description \\
    \hline
    \hline
    $12$   & simulation parameters \\
    $1000$ & nLTT statistic \\
    \hline
    $12$ & simulation parameters \\
    $1$ & expected duration of speciation (from Etienne \& Rosindell 2012) \\
    $1$ & number of taxa in incipient species tree (also: number of subspecies) \\
    $1$ & number of taxa in species tree (also: number of species) \\
    $1$ & mean nLTT statistic \\
    $1$ & median nLTT statistic \\
    $1$ & lower bound of 95\% HPD value of nLTT statistic \\
    $1$ & upper bound of 95\% HPD value of nLTT statistic \\
    $11$ & ESSes of all parameters estimated by BEAST2 (see specs below) \\
    \hline
  \end{tabular}
  \caption{
    Specification of the data sets. $n$ denotes the number
    of columns a certain item will occupy. Above the horizontal line,
    the content of the raw data file is specified, resulting in a table of 
    1012 columns and 40K rows. Below the horizontal line,
    the content of the summary data file specified, resulting in a table of
    42 columns and 40K rows.
  }
  \label{table:raw_data_set_specs}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\usepackage[english]{babel}

@@

@@

The model prior most relevant here, is the tree prior. The tree prior
embodies the assumed speciation model. Each speciation model has a
different expected branch length distribution. It has been shown 
[NOTE: where?] that the effect of the prior on the posterior distribution 
of trees can be substantial.

@@

There are problems with ignoring protracted speciation. A very pragmatic
example is the conservation of an underestimated amount of species. 
But from a more theoretical perspective, the
observed decline of speciation rate in time (excluding the contemporary human-driven
mass extinction event) may be explained by these new species not being 
recognized yet. Another problem is that the most recent common 
ancestor (MRCA) of a young species duo is estimated 
too close to the present, as in the BD model a species can be 
ultra-young (as speciation is assumed to be instantaneous), 
where the PBD model will place these MRCAs
further back in time (as two events need to have finished completion).

From a theoretical point of view, there is another reason why protracted
speciation has not yet become a member of our toolbox: there is no framework
where it fits in: a framework assumes either an analysis at the species or
subspecies level. When picking a tool that does its analysis at the
species level, there is no protocol how to select which incipient species
gets to represent the full species it is still a member of. When picking
a tool at the sub-species level, the incipient state disappears as a
sub-species lineage, and the process degrades to an ordinary birth-death 
process.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l}
    \hline
    $n$ & Description \\
    \hline
    \hline
    $12$   & simulation parameters \\
    $1000$ & nLTT statistic \\
    \hline
    $12$ & simulation parameters \\
    $1$ & expected duration of speciation (from Etienne \& Rosindell 2012) \\
    $1$ & number of taxa in incipient species tree (also: number of subspecies) \\
    $1$ & number of taxa in species tree (also: number of species) \\
    $1$ & mean nLTT statistic \\
    $1$ & median nLTT statistic \\
    $1$ & lower bound of 95\% HPD value of nLTT statistic \\
    $1$ & upper bound of 95\% HPD value of nLTT statistic \\
    $11$ & ESSes of all parameters estimated by BEAST2 (see specs below) \\
    \hline
  \end{tabular}
  \caption{
    Specification of the data sets. $n$ denotes the number
    of columns a certain item will occupy. Above the horizontal line,
    the content of the raw data file is specified, resulting in a table of 
    1012 columns and 40K rows. Below the horizontal line,
    the content of the summary data file specified, resulting in a table of
    40 columns and 40K rows.
  }
  \label{table:raw_data_set_specs}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
